Relationships in the extremes and their influence on competition and coexistence

### 2022-01-20 paper/Paper.Rnw
writing discussion: paragraph 3: future work in changign ATAs

\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence of absence of ATA 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – the IGR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values ATA exclusion occurs where ATAs are the 
driving mechanism of the failure of coexistence and the lack of ATAs would result in 
a positive IGR. 
Subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists}.
The parameter subspaces of ATA rescue and ATA exclusion suggest that some species may be ATA specialists  - only persisting in habitat patches where ATAs contribute to its coexistence. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

\underline{future work: study ATAs in environmental variables and how theyre changing - tie in with specialist}
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

%Although environmental fluctuations is helpful for coexistence and storage effect, unusual extreme fluctuations can be a hinderance on coexistence by introducing or changing ATAs between environmental variables. 
Consistent with the mechanisms of biodiversity maintenance and specifically 
fluctuation-dependent mechanisms (cite chesson), higher values of environmental fluctuation 
resulted in higher IGR’s. Both of our new terms, $\Delta_i^{[EC]}$ and $\Delta_i^{[E \nat C]}$ 
also increased in magnitude with increased environmental fluctuation. Increased environmental 
fluctuation is reflected in increased $\sigma$ in the lottery model and increased $a$ in the 
competitive diatom model.   mean temp and period?

Distinct increase in contribution of ATAs to coexistence as environmental fluctuations increase may suggest that ATAs may become more important to consider for coexistence and possibly other ecological phenomena as the environment becomes more variable, a consequence of climate change and extreme climatic events.

This may suggest that, if environmental fluctuations become more variable, 
the effect of ATAs becomes more important and non-negligible (rescue or exclusion). 
{Might be interesting to study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.} 
%This would be a whole potential 
%research program which presumably would start with a more detailed study of how tail associations of environmental
%variables may be changing under global change.

@@

\underline{the new effects are not "weak noise" effects}
if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.
From proposal: 
Classical analytic derivations of MCT using Taylor expansion approximations completely miss out
on any additions to coexistence beyond simple correlations. 
In fact, storage effects in analytic MCT is computationally equal to my new term, Δ^([E||C]), 
the contributions of correlation per se. In newer, simulation based MCT, storage effect is 
the sum of my two new terms, Δ^((EC))=Δ^([E||C])+Δ^([EC]), obscuring the effects of ATAs

>>>>

\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence of absence of ATA 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – the IGR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values ATA exclusion occurs where ATAs are the 
driving mechanism of the failure of coexistence and the lack of ATAs would result in 
a positive IGR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

\underline{future work: study ATAs in environmental variables and how theyre changing - tie in with specialist}

Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
Distinct increases in the magnitude of ATA contribution with increasing
environmental fluctuations, as shown in our results and
figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns of environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

%Although environmental fluctuations is helpful for coexistence and storage effect, unusual extreme fluctuations can be a hinderance on coexistence by introducing or changing ATAs between environmental variables. 
%onsistent with the mechanisms of biodiversity maintenance and specifically 
%luctuation-dependent mechanisms (cite chesson), higher values of environmental fluctuation 
%esulted in higher IGR’s. Both of our new terms, $\Delta_i^{[EC]}$ and $\Delta_i^{[E \nat C]}$ 
%lso increased in magnitude with increased environmental fluctuation. Increased environmental 
%luctuation is reflected in increased $\sigma$ in the lottery model and increased $a$ in the 
%ompetitive diatom model.   mean temp and period?

%This would be a whole potential 
%research program which presumably would start with a more detailed study of how tail associations of environmental
%variables may be changing under global change.

@@

\underline{the new effects are not "weak noise" effects}
if sigma really small, ATA would not matter?
%go back to emails
%Maybe have a paragraph about how Chesson's weak noise and other assumptions make it impossible to detect this new effect
%via that framework. It's only the less-assumption-reliant simulation approach of Ellner et al that makes it possible to 
%study this new term in the decomp. Need to study a bit to substantiate this properly, or get Steve Ellner help.

From proposal:

Classical analytic derivations of MCT using Taylor expansion approximations completely miss out
on any additions to coexistence beyond simple correlations. 
In fact, storage effects in analytic MCT is computationally equal to $\Delta^{[E\natC]}$, 
the contributions of correlation per se. In newer, simulation based MCT, storage effect is 
the sum of the two new terms, $\Delta^{(EC)}=\Delta^{[E\natC]}+\Delta^{[EC]}$, obscuring the effects of ATAs.
